FOODiE, Fortran Object oriented Ordinary Differential Equations integration library based on Abstract Calculus Pattern

### 2015-10-06 papers/announcement/manuscript.tex
Improve manuscript
Improve manuscript:

+ complete API review and disccusion;
+ almost complete Oscillation test disccusion (must be completed the
results analysis).

The prototype of IVP can be expressed as:
\begin{equation}
  \begin{matrix}
  U_t = R(t,U) \\
  U_0 = U(t_0)
  \end{matrix}
\label{eq:IVP}
\end{equation}
where $U(t)$ is the vector of state variables being a function of the time-like independent variable $t$, $U_t = \frac{dU}{dt} = R(t,U)$ is the (vectorial) residual function and $U(t_0)$ is the (vectorial) initial conditions, namely the state variables function evaluated at the initial time $t_0$. In general, the residual function $R$ is a function of the state variable $U$ through which it is a function of time, but it can be also a direct function of time, thus in general $R=R(t,U(t))$ holds.

@@

It is worth noting that the state vector variables $U$ and its corresponding residual function $U_t = \frac{dU}{dt} = R(t,U)$ are \emph{problem dependent}: the number and the meaning of the state variables as well as the equations governing their evolution (that are embedded into the residual function) are different for the Navier-Stokes conservation laws with respect the Burgers one, as an example. Nevertheless, the \emph{solvers} used for the prediction of the Navier-Stokes equations evolution (hereafter the \emph{solution}) are the same that are used for Burgers equations time-integration. As a consequence, the solution of the IVP model prototype can be generalized, allowing the application of the same solver to many different problems, thus eliminating the necessity to re-implement the same solver for each different problem.

@@

In many (most) circumstances, the solution of equation \ref{eq:IVP} cannot be computed in a closed, exact form (even if it exists and is unique) due to the complexity and nature of the residual functions, that is often non linear. Consequently, the problem is often solved relying on a numerical approach: the solution of system \ref{eq:IVP} at a time $t^n$, namely $U(t^n)$, is approximated by a subsequent time-marching approximations $U_0=u_0 \rightarrow u_1 \rightarrow u_2 \rightarrow ... \rightarrow u_N\approx U(t^n)$ where the relation $u_i \rightarrow u_{i+1}$ implies a \emph{stepping, numerical integration} from the time $t^i$ to time $t^{i+1}$ and $N$ is the total number of numerical time steps necessary to evolve the initial conditions toward the searched solution $U(t^n)$. To this aim, many numerical schemes have been devised. Notably, the numerical schemes of practical usefulness must posses some necessary proprieties such as \emph{consistency} and \emph{stability} to ensure that the numerical approximation \emph{converges} to the exact solution as the numerical time step tends to zero. A detailed discussion of these details is out the scope of the present work and is omitted. Here, we briefly recall some classifications necessary to introduce the schemes implemented into the FOODiE library.

@@

The equations \ref{eq:RK} and \ref{eq:RK-stage} show that Runge-Kutta methods do not require any additional differentiations of the ODE system for achieving high order accuracy, rather they require additional evaluations of the residual function $R$.

@@

Adams-Bashforth methods belong to the more general (linear) multi-step family of schemes. This kind of schemes has been designed to achieve a more accurate solution than the $1^{st}$ Euler scheme using the information coming from the solutions already computed at previous time steps. Typically only one new residual function $R$ evaluation is required at each time step, whereas Runge-Kutta schemes require many of them.

@@

\subsubsection{Time derivative procedure, the residual function}

@@

This procedure-function takes three arguments, the first passed as a \emph{type bounded} argument, while the latter are optionals, and it returns an integrand object. The passed dummy argument, \emph{self}, is a polymorphic argument that could be any extensions of the \emph{integrand} ADT. The optional argument \emph{n} indicates the \emph{time level} at which the residual function must be evaluated: this argument is necessary for multi-step (often referred as multi-level) solvers such as the Adams-Bashforth or leapfrog ones, but it can be omitted for other classes of solvers such the one-step multi-stages family. The optional argument \emph{t} is the \emph{time} at which the residual function must be computed: it can be omitted in the case the residual function does not depend directly on time.

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={implementation of the \emph{integrate} method of Euler solver},label={list:euler_integrate}]
subroutine integrate(U, Dt, t)
class(integrand),    intent(INOUT) :: U  !< Field to be integrated.
real(R_P),           intent(IN)    :: Dt !< Time step.
real(R_P), optional, intent(IN)    :: t  !< Time.
U = U + U%t(t=t) * Dt
return
endsubroutine integrate
\end{lstlisting}
This method takes three arguments, the first argument is an integrand class, it being the integrand field that must integrated one-step-over in time, the second is the time step used and the third, that is optional, is the current time value that is passed to the residual function for taking into account the cases where the time derivative explicitly depends on time. The time step is not automatically computed (for example inspecting the passed integrand field), thus its value must be externally computed and passed to the \emph{integrate} method.

As shown from listing \ref{list:euler_integrate} the \emph{integrand} ADT has allowed a very high-level implementation of the Euler scheme. For example let us compare the implementation instructions with the Euler's scheme equation \ref{eq:solver-euler}:

$$
U\left(t^{n+1}\right) = U\left(t^n\right) +\Delta t \cdot R\left[t^n, U\left(t^n\right)\right] \rightarrow U = U + U\%t(t=t) * Dt
$$

As shown, the Fortran implementation is almost equivalent to the rigorous mathematical formulation. This aspect directly implies that the implementation of Euler's scheme (as of all other schemes) is very clear, concise and less-errors-prone than an \emph{hard-coded} implementation where the Euler's solver must be implemented for each specific definition of the integrand $U$.

@@

\begin{itemize}
  \item \emph{init}: initialize the integrator among the possibilities offered by the class of solvers;
  \item \emph{destroy}: destroy the integrator previously initialized, eventually freeing the allocated dynamic memory registers;
  \item \emph{integrate}: integrate integrand field one-step-over in time.
  \end{itemize}

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of \emph{tvd\_runge\_kutta\_integrator} type},label={list:tvd_rk_solver}]
type :: tvd_runge_kutta_integrator
  integer(I_P)           :: stages=0  ! Number of stages.
  real(R_P), allocatable :: alph(:,:) ! alpha Butcher's coefficients.
  real(R_P), allocatable :: beta(:)   ! beta Butcher's coefficients.
  real(R_P), allocatable :: gamm(:)   ! gamma Butcher's coefficients.
  contains
    procedure, pass(self), public :: destroy   ! Destroy the integrator.
    procedure, pass(self), public :: init      ! Initialize (create) the integrator.
    procedure, pass(self), public :: integrate ! Integrate integrand field.
    final                         :: finalize  ! Finalize object.
endtype tvd_runge_kutta_integrator
\end{lstlisting}

After the solver has been initialized it can used for integrating an integrand field, as shown in listing \ref{list:use_tvd_rk_solver}.

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={implementation of the \emph{integrate} method of TVD/SSP Runge-Kutta class},label={list:tvd_rk_integrate}]
subroutine integrate(self, U, stage, Dt, t)
class(tvd_runge_kutta_integrator), intent(IN)    :: self      !< Actual RK integrator.
class(integrand),                  intent(INOUT) :: U         !< Field to be integrated.
class(integrand),                  intent(INOUT) :: stage(1:) !< Runge-Kutta stages [1:stages].
real(R_P),                         intent(IN)    :: Dt        !< Time step.
real(R_P),                         intent(IN)    :: t         !< Time.
integer(I_P)                                     :: s         !< First stages counter.
integer(I_P)                                     :: ss        !< Second stages counter.
select type(stage)
class is(integrand)
  do s=1, self%stages
    stage(s) = U
    do ss=1, s - 1
      stage(s) = stage(s) + stage(ss) * (Dt * self%alph(s, ss))
    enddo
    stage(s) = stage(s)%t(t=t + self%gamm(s) * Dt)
  enddo
  do s=1, self%stages
    U = U +  stage(s) * (Dt * self%beta(s))
  enddo
endselect
return
endsubroutine integrate
\end{lstlisting}

This method takes five arguments, the first argument is passed as bounded argument and it is the solver itself, the second is of an integrand class, it being the integrand field that must integrated one-step-over in time, the third is the stages array for storing the stages computations, the fourth is the time step used and the fifth, that is optional, is the current time value that is passed to the residual function for taking into account the cases where the time derivative explicitly depends on time. The time step is not automatically computed (for example inspecting the passed integrand field), thus its value must be externally computed and passed to the \emph{integrate} method.

@@

As it happens for the Euler's scheme, the \emph{integrand} ADT greatly simplifies the Runge-Kutta algorithm implementation. For example let us compare the implementation instructions with the Runge-Kutta equations \ref{eq:RK} and \ref{eq:RK-stage}:

$$
\begin{matrix}
  K_s = R\left(t^n + \gamma_s \Delta t, U^n+\Delta t\sum\limits_{l=1}^{s-1}{\alpha_{s,l} K_l} \right) \\
  U^{n+1} = U^{n} + \Delta t \cdot\sum\limits_{s=1}^{N_{s}}{\beta_s K_s}
\end{matrix}
\rightarrow
\begin{matrix}
  stage(s) = stage(s) + stage(ss) * (Dt * self\%alph(s, ss)) \\
  U = U +  stage(s) * (Dt * self\%beta(s))
\end{matrix}
$$
where the syntax "$(s)$" and "$(ss)$" imply the summation operation. As for the Euler's scheme, the Fortran implementation is almost equivalent to the rigorous mathematical formulation.

@@

@@

@@

@@

{\color{red} To be completed.}

\begin{equation}
\begin{matrix}
U_t = R(U)  \\
U = \begin{bmatrix}
x \\
y
\end{bmatrix}\;\;\;
R(U) = \begin{bmatrix}
-f y \\
f x
\end{bmatrix}
\end{matrix}
\label{eq:oscillation}
\end{equation}

where the frequency is chosen as $f=10^4$. The ODE system \ref{eq:oscillation} is completed by the following initial conditions:

\begin{equation}
\begin{matrix}
  x(t_0) = 0 \\
  y(t_0) = 1
\end{matrix}
\label{eq:oscillation-ic}
\end{equation}

where $t_0=0$ is the initial time considered. The exact solution is:

\begin{equation}
\begin{matrix}
  x(t_0 + \Delta t) = X_0 cos(f \Delta t) - y_0 sin(f \Delta t) \\
  y(t_0 + \Delta t) = X_0 sin(f \Delta t) + y_0 cos(f \Delta t)
\end{matrix}
\label{eq:oscillation-exact}
\end{equation}

where $\Delta t$ is an arbitrary time step.

\subsubsection{Errors Analysis}

For the analysis of the accuracy of each solvers implemented into FOODiE library, we have integrated the Oscillation equations \ref{eq:oscillation} with different, decreasing time steps in the range $[5000, 2500, 1250, 625, 320, 100]$.

>>>>

The prototype of IVP can be expressed as:
\begin{equation}
  \begin{matrix}
  U_t = R(t,U) \\
  U_0 = U(t_0)
  \end{matrix}
\label{eq:IVP}
\end{equation}
where $U(t)$ is the vector of state variables being a function of the time-like independent variable $t$, $U_t = \frac{dU}{dt} = R(t,U)$ is the (vectorial) residuals function and $U(t_0)$ is the (vectorial) initial conditions, namely the state variables function evaluated at the initial time $t_0$. In general, the residuals function $R$ is a function of the state variable $U$ through which it is a function of time, but it can be also a direct function of time, thus in general $R=R(t,U(t))$ holds.

@@

It is worth noting that the state vector variables $U$ and its corresponding residuals function $U_t = \frac{dU}{dt} = R(t,U)$ are \emph{problem dependent}: the number and the meaning of the state variables as well as the equations governing their evolution (that are embedded into the residuals function) are different for the Navier-Stokes conservation laws with respect the Burgers one, as an example. Nevertheless, the \emph{solvers} used for the prediction of the Navier-Stokes equations evolution (hereafter the \emph{solution}) are the same that are used for Burgers equations time-integration. As a consequence, the solution of the IVP model prototype can be generalized, allowing the application of the same solver to many different problems, thus eliminating the necessity to re-implement the same solver for each different problem.

@@

In many (most) circumstances, the solution of equation \ref{eq:IVP} cannot be computed in a closed, exact form (even if it exists and is unique) due to the complexity and nature of the residuals functions, that is often non linear. Consequently, the problem is often solved relying on a numerical approach: the solution of system \ref{eq:IVP} at a time $t^n$, namely $U(t^n)$, is approximated by a subsequent time-marching approximations $U_0=u_0 \rightarrow u_1 \rightarrow u_2 \rightarrow ... \rightarrow u_N\approx U(t^n)$ where the relation $u_i \rightarrow u_{i+1}$ implies a \emph{stepping, numerical integration} from the time $t^i$ to time $t^{i+1}$ and $N$ is the total number of numerical time steps necessary to evolve the initial conditions toward the searched solution $U(t^n)$. To this aim, many numerical schemes have been devised. Notably, the numerical schemes of practical usefulness must posses some necessary proprieties such as \emph{consistency} and \emph{stability} to ensure that the numerical approximation \emph{converges} to the exact solution as the numerical time step tends to zero. A detailed discussion of these details is out the scope of the present work and is omitted. Here, we briefly recall some classifications necessary to introduce the schemes implemented into the FOODiE library.

@@

The equations \ref{eq:RK} and \ref{eq:RK-stage} show that Runge-Kutta methods do not require any additional differentiations of the ODE system for achieving high order accuracy, rather they require additional evaluations of the residuals function $R$.

@@

Adams-Bashforth methods belong to the more general (linear) multi-step family of schemes. This kind of schemes has been designed to achieve a more accurate solution than the $1^{st}$ Euler scheme using the information coming from the solutions already computed at previous time steps. Typically only one new residuals function $R$ evaluation is required at each time step, whereas Runge-Kutta schemes require many of them.

@@

\subsubsection{Time derivative procedure, the residuals function}

@@

This procedure-function takes three arguments, the first passed as a \emph{type bounded} argument, while the latter are optionals, and it returns an integrand object. The passed dummy argument, \emph{self}, is a polymorphic argument that could be any extensions of the \emph{integrand} ADT. The optional argument \emph{n} indicates the \emph{time level} at which the residuals function must be evaluated: this argument is necessary for multi-step (often referred as multi-level) solvers such as the Adams-Bashforth or leapfrog ones, but it can be omitted for other classes of solvers such the one-step multi-stages family. The optional argument \emph{t} is the \emph{time} at which the residuals function must be computed: it can be omitted in the case the residuals function does not depend directly on time.

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={implementation of the \emph{integrate} method of Euler solver},label={list:euler_integrate}]
subroutine integrate(U, Dt, t)
class(integrand),    intent(INOUT) :: U  !< Field to be integrated.
real(R_P),           intent(IN)    :: Dt !< Time step.
real(R_P), optional, intent(IN)    :: t  !< Time.
U = U + U%t(t=t) * Dt
return
endsubroutine integrate
\end{lstlisting}
This method takes three arguments, the first argument is an integrand class, it being the integrand field that must integrated one-step-over in time, the second is the time step used and the third, that is optional, is the current time value that is passed to the residuals function for taking into account the cases where the time derivative explicitly depends on time. The time step is not automatically computed (for example inspecting the passed integrand field), thus its value must be externally computed and passed to the \emph{integrate} method.

@@

\begin{itemize}
  \item \emph{init}: initialize the integrator accordingly the possibilities offered by the class of solvers;
  \item \emph{destroy}: destroy the integrator previously initialized, eventually freeing the allocated dynamic memory registers;
  \item \emph{integrate}: integrate integrand field one-step-over in time.
  \end{itemize}

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of \emph{tvd\_runge\_kutta\_integrator} type},label={list:tvd_rk_solver}]
type :: tvd_runge_kutta_integrator
  integer(I_P)           :: stages=0  ! Number of stages.
  real(R_P), allocatable :: alph(:,:) ! alpha Butcher's coefficients.
  real(R_P), allocatable :: beta(:)   ! beta Butcher's coefficients.
  real(R_P), allocatable :: gamm(:)   ! gamma Butcher's coefficients.
  contains
    procedure, pass(self), public :: destroy
    procedure, pass(self), public :: init
    procedure, pass(self), public :: integrate
    final                         :: finalize
endtype tvd_runge_kutta_integrator
\end{lstlisting}

After the solver has been initialized it can be used for integrating an integrand field, as shown in listing \ref{list:use_tvd_rk_solver}.

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={implementation of the \emph{integrate} method of explicit TVD/SSP Runge-Kutta class},label={list:tvd_rk_integrate}]
subroutine integrate(self, U, stage, Dt, t)
class(tvd_runge_kutta_integrator), intent(IN)    :: self      ! Actual RK integrator.
class(integrand),                  intent(INOUT) :: U         ! Field to be integrated.
class(integrand),                  intent(INOUT) :: stage(1:) ! Runge-Kutta stages [1:stages].
real(R_P),                         intent(IN)    :: Dt        ! Time step.
real(R_P),                         intent(IN)    :: t         ! Time.
integer(I_P)                                     :: s         ! First stages counter.
integer(I_P)                                     :: ss        ! Second stages counter.
select type(stage)
class is(integrand)
  do s=1, self%stages
    stage(s) = U
    do ss=1, s - 1
      stage(s) = stage(s) + stage(ss) * (Dt * self%alph(s, ss))
    enddo
    stage(s) = stage(s)%t(t=t + self%gamm(s) * Dt)
  enddo
  do s=1, self%stages
    U = U +  stage(s) * (Dt * self%beta(s))
  enddo
endselect
return
endsubroutine integrate
\end{lstlisting}

This method takes five arguments, the first argument is passed as bounded argument and it is the solver itself, the second is of an integrand class, it being the integrand field that must integrated one-step-over in time, the third is the stages array for storing the stages computations, the fourth is the time step used and the fifth, that is optional, is the current time value that is passed to the residuals function for taking into account the cases where the time derivative explicitly depends on time. The time step is not automatically computed (for example inspecting the passed integrand field), thus its value must be externally computed and passed to the \emph{integrate} method.

@@

@@

The low storage variant of Runge-Kutta class of solvers is exposed as a single derived type named \emph{ls\_runge\_kutta\_integrator}. This type provides three methods:

\begin{itemize}
  \item \emph{init}: initialize the integrator accordingly the possibilities offered by the class of solvers;
  \item \emph{destroy}: destroy the integrator previously initialized, eventually freeing the allocated dynamic memory registers;
  \item \emph{integrate}: integrate integrand field one-step-over in time.
  \end{itemize}

As common for FOODiE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODiE integrator type, see listing \ref{list:define_ls_rk_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an explicit low storage Runge-Kutta integrator},label={list:define_ls_rk_solver}]
use FOODiE, only: ls_runge_kutta_integrator
type(ls_runge_kutta_integrator) :: integrator
\end{lstlisting}

Once an integrator of this type has been instantiated, it must be initialized before used, for example see listing \ref{list:init_ls_rk_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={example of initialization of an explicit low storage Runge-Kutta integrator},label={list:init_ls_rk_solver}]
call integrator%init(stages=5)
\end{lstlisting}

In the listing \ref{list:init_ls_rk_solver} a 5-stages solver has been initialized. As a matter of facts, from the equation \ref{eq:RK-ls} a solver belonging to this class is completely defined once the number of stages adopted has been chosen. The complete definition of the \emph{ls\_runge\_kutta\_integrator} type is reported into listing \ref{list:ls_rk_solver}. As shown, the Williamson's coefficients are stored as allocatable arrays the values of which are initialized by the \emph{init} method.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of \emph{ls\_runge\_kutta\_integrator} type},label={list:ls_rk_solver}]
type :: ls_runge_kutta_integrator
  integer(I_P)           :: stages=0 ! Number of stages.
  real(R_P), allocatable :: A(:)     ! Low storage *A* coefficients.
  real(R_P), allocatable :: B(:)     ! Low storage *B* coefficients.
  real(R_P), allocatable :: C(:)     ! Low storage *C* coefficients.
  contains
    procedure, pass(self), public :: destroy
    procedure, pass(self), public :: init
    procedure, pass(self), public :: integrate
    final                         :: finalize
endtype ls_runge_kutta_integrator
\end{lstlisting}

After the solver has been initialized it can be used for integrating an integrand field, as shown in listing \ref{list:use_ls_rk_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={example of usage of a low storage Runge-Kutta integrator},label={list:use_ls_rk_solver}]
type(my_integrand) :: my_field
type(my_integrand) :: my_stages(1:2)
call integrator%integrate(U=my_field, stage=my_stage, Dt=0.1)
\end{lstlisting}
where \emph{my\_integrand} is a concrete (valid) extension of \emph{integrand} ADT. Listing \ref{list:use_ls_rk_solver} shows that the memory registers necessary for storing the Runge-Kutta stages must be supplied by the client code, as it happens of the TVD/SSP Runge-Kutta class. However, now the registers necessary is always 2, independently on the number of stages used, that in the example considered are 5.

The complete implementation of the \emph{integrate} method of the explicit low storage Runge-Kutta class of solvers is reported in listing \ref{list:ls_rk_integrate}.

\begin{lstlisting}[firstnumber=1,style=code,caption={implementation of the \emph{integrate} method of explicit low storage Runge-Kutta class},label={list:ls_rk_integrate}]
subroutine integrate(self, U, stage, Dt, t)
class(ls_runge_kutta_integrator), intent(IN)    :: self       ! Actual RK integrator.
class(integrand),                 intent(INOUT) :: U          ! Field to be integrated.
class(integrand),                 intent(INOUT) :: stage(1:2) ! Runge-Kutta registers [1:2].
real(R_P),                        intent(IN)    :: Dt         ! Time step.
real(R_P),                        intent(IN)    :: t          ! Time.
integer(I_P)                                    :: s          ! First stages counter.
select type(stage)
class is(integrand)
  stage(1) = U
  stage(2) = U*0._R_P
  do s=1, self%stages
    stage(2) = stage(2) * self%A(s) + stage(1)%t(t=t + self%C(s) * Dt) * Dt
    stage(1) = stage(1) + stage(2) * self%B(s)
  enddo
  U = stage(1)
endselect
return
endsubroutine integrate
\end{lstlisting}

This method takes five arguments, the first argument is passed as bounded argument and it is the solver itself, the second is of an integrand class, it being the integrand field that must integrated one-step-over in time, the third is the stages array for storing the stages computations, the fourth is the time step used and the fifth, that is optional, is the current time value that is passed to the residuals function for taking into account the cases where the time derivative explicitly depends on time. The time step is not automatically computed (for example inspecting the passed integrand field), thus its value must be externally computed and passed to the \emph{integrate} method.

@@

The explicit Adams-Bashforth class of solvers is exposed as a single derived type named \emph{adams\_bashforth\_integrator}. This type provides three methods:

\begin{itemize}
  \item \emph{init}: initialize the integrator accordingly the possibilities offered by the class of solvers;
  \item \emph{destroy}: destroy the integrator previously initialized, eventually freeing the allocated dynamic memory registers;
  \item \emph{integrate}: integrate integrand field one-step-over in time.
  \end{itemize}

As common for FOODiE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODiE integrator type, see listing \ref{list:define_ab_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an explicit Adams-Bashforth integrator},label={list:define_ab_solver}]
use FOODiE, only: adams_bashforth_integrator
type(adams_bashforth_integrator) :: integrator
\end{lstlisting}

Once an integrator of this type has been instantiated, it must be initialized before used, for example see listing \ref{list:init_ab_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={example of initialization of an explicit Adams-Bashforth integrator},label={list:init_ab_solver}]
call integrator%init(steps=4)
\end{lstlisting}

In the listing \ref{list:init_ab_solver} a 4-steps solver has been initialized. As a matter of facts, from the equation \ref{eq:AB} a solver belonging to this class is completely defined once the number of time steps adopted has been chosen. The complete definition of the \emph{adams\_bashforth\_integrator} type is reported into listing \ref{list:ab_solver}. As shown, the linear coefficients are stored as allocatable arrays the values of which are initialized by the \emph{init} method.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of \emph{adams\_bashforth\_integrator} type},label={list:ab_solver}]
type :: adams_bashforth_integrator
  private
  integer(I_P)           :: steps=0 ! Number of time steps.
  real(R_P), allocatable :: b(:)    ! \(b\) coefficients.
  contains
    procedure, pass(self), public :: destroy
    procedure, pass(self), public :: init
    procedure, pass(self), public :: integrate
    final                         :: finalize
endtype adams_bashforth_integrator
\end{lstlisting}

After the solver has been initialized it can be used for integrating an integrand field, as shown in listing \ref{list:use_ab_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={example of usage of an Adams-Bashforth integrator},label={list:use_ab_solver}]
real               :: times(1:4)
type(my_integrand) :: my_field
call integrator%integrate(U=my_field, Dt=Dt, t=times)
\end{lstlisting}
where \emph{my\_integrand} is a concrete (valid) extension of \emph{integrand} ADT and \emph{times} are the time at each 4 steps considered for the current one-step-over integration.

The complete implementation of the \emph{integrate} method of the explicit Adams-Bashforth class of solvers is reported in listing \ref{list:ab_integrate}.

\begin{lstlisting}[firstnumber=1,style=code,caption={implementation of the \emph{integrate} method of explicit Adams-Bashforth class},label={list:ab_integrate}]
subroutine integrate(self, U, Dt, t)
class(adams_bashforth_integrator), intent(IN)    :: self ! Actual AB integrator.
class(integrand),                  intent(INOUT) :: U    ! Field to be integrated.
real(R_P),                         intent(IN)    :: Dt   ! Time steps.
real(R_P),                         intent(IN)    :: t(:) ! Times.
integer(I_P)                                     :: s    ! Steps counter.
do s=1, self%steps
  U = U + U%t(n=s, t=t(s)) * (Dt * self%b(s))
enddo
call U%update_previous_steps
return
endsubroutine integrate
\end{lstlisting}

This method takes four arguments, the first argument is passed as bounded argument and it is the solver itself, the second is of an integrand class, it being the integrand field that must integrated one-step-over in time, the third is the time step used and the fourth is an array of the time values of the steps considered for the current one-step-over integration that are passed to the residuals function for taking into account the cases where the time derivative explicitly depends on time. The time step is not automatically computed (for example inspecting the passed integrand field), thus its value must be externally computed and passed to the \emph{integrate} method.

It is worth noting that there is an explicit call to the method \emph{update\_previous\_steps} for updating the values of the previously computed steps.

@@

The explicit Leapfrog class of solvers is exposed as a single derived type named \emph{leapfrog\_integrator}. This type provides three methods:

\begin{itemize}
  \item \emph{init}: initialize the integrator accordingly the possibilities offered by the class of solvers;
  \item \emph{integrate}: integrate integrand field one-step-over in time.
  \end{itemize}

As common for FOODiE solvers, for using such a solver it must be previously defined as an instance of the exposed FOODiE integrator type, see listing \ref{list:define_lf_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of an explicit Leapfrog integrator},label={list:define_lf_solver}]
use FOODiE, only: leapfrog_integrator
type(leapfrog_integrator) :: integrator
\end{lstlisting}

Once an integrator of this type has been instantiated, it must be initialized before used, for example see listing \ref{list:init_lf_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={example of initialization of an explicit Leapfrog integrator},label={list:init_lf_solver}]
! default coefficients nu=0.01, alpha=0.5
call integrator%init()
! custom coefficients
call integrator%init(nu=0.015, alpha=0.6)
\end{lstlisting}

The complete definition of the \emph{leapfrog\_integrator} type is reported into listing \ref{list:lf_solver}. As shown, the filter coefficients are initialized to zero, suitable values are initialized by the \emph{init} method.

\begin{lstlisting}[firstnumber=1,style=code,caption={definition of \emph{leapfrog\_integrator} type},label={list:lf_solver}]
type :: leapfrog_integrator
  private
  real(R_P) :: nu=0.01_R_P   ! Robert-Asselin filter coefficient.
  real(R_P) :: alpha=0.5_R_P ! Robert-Asselin-Williams filter coefficient.
  contains
    procedure, pass(self), public :: init
    procedure, pass(self), public :: integrate
endtype leapfrog_integrator
\end{lstlisting}

After the solver has been initialized it can be used for integrating an integrand field, as shown in listing \ref{list:use_lf_solver}.

\begin{lstlisting}[firstnumber=1,style=code,caption={example of usage of a Leapfrog integrator},label={list:use_lf_solver}]
real               :: times(1:2)
type(my_integrand) :: filter_displacement
type(my_integrand) :: my_field
call integrator%integrate(U=my_field, filter=filter_displacement, Dt=Dt, t=times)
\end{lstlisting}
where \emph{my\_integrand} is a concrete (valid) extension of \emph{integrand} ADT, \emph{filter\_displacement} is the register necessary for computing the eventual displacement of the applied filter and \emph{times} are the time at each 2 steps considered for the current one-step-over integration.

The complete implementation of the \emph{integrate} method of the explicit Leapfrog class of solvers is reported in listing \ref{list:lf_integrate}.

\begin{lstlisting}[firstnumber=1,style=code,caption={implementation of the \emph{integrate} method of explicit Leapfrog class},label={list:lf_integrate}]
subroutine integrate(self, U, filter, Dt, t)
class(leapfrog_integrator), intent(IN)    :: self   ! LF integrator.
class(integrand),           intent(INOUT) :: U      ! Field to be integrated.
class(integrand), optional, intent(INOUT) :: filter ! Filter field displacement.
real(R_P),                  intent(in)    :: Dt     ! Time step.
real(R_P),                  intent(IN)    :: t      ! Time.
U = U%previous_step(n=1) + U%t(n=2, t=t) * (Dt * 2._R_P)
if (present(filter)) then
  filter = (U%previous_step(n=1) - U%previous_step(n=2) * 2._R_P + U) * self%nu * 0.5_R_P
endif
call U%update_previous_steps(filter=filter, weights=[self%alpha, self%alpha - 1._R_P])
return
endsubroutine integrate
\end{lstlisting}

This method takes five arguments, the first argument is passed as bounded argument and it is the solver itself, the second is of an integrand class, it being the integrand field that must integrated one-step-over in time, the third is the optional filter-displacement-register, the fourth is the time step used and the fifth is an array of the time values of the steps considered for the current one-step-over integration that are passed to the residuals function for taking into account the cases where the time derivative explicitly depends on time. The time step is not automatically computed (for example inspecting the passed integrand field), thus its value must be externally computed and passed to the \emph{integrate} method. It is worth noting that if the filter displacement argument is not passed, the solver reverts back to the standard unfiltered Leapfrog method.

It is worth noting that there is an explicit call to the method \emph{update\_previous\_steps} for updating the values of the previously computed steps.

\subsection{General Remarks}\label{subsec:API-remarks}

Table \ref{tab:foodie-vs-math} presents a comparison of the relevant parts of equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:RK-stage}, \ref{eq:RK-ls}, \ref{eq:AB} and \ref{eq:leapfrog} with the corresponding FOODiE implementations reported in listings \ref{list:euler_integrate}, \ref{list:tvd_rk_integrate}, \ref{list:ls_rk_integrate}, \ref{list:ab_integrate} and \ref{list:lf_integrate}, respectively. This comparison proves that the \emph{integrand} ADT has allowed a very high-level implementation syntax. The Fortran implementation is almost equivalent to the rigorous mathematical formulation. This aspect directly implies that the implementation of a ODE solver into the FOODiE library is very clear, concise and less-errors-prone than an \emph{hard-coded} implementation where the solvers must be implemented for each specific definition of the integrand type, it being not abstract.

\begin{table}[!ht]
  \centering
  \caption{Comparison between rigorous mathematical formulation and FOODiE high-level implementation; the syntax "$(s)$" and "$(ss)$" imply the summation operation}\label{tab:foodie-vs-math}
  \resizebox{0.99\textwidth}{!}{%
  \begin{tabular}{c|cc}
    \toprule
    {\sc Solver} & {\sc Mathematical formulation} & {\sc FOODiE implementation} \\
    \hline
    explicit forward Euler & $U\left(t^{n+1}\right) = U\left(t^n\right) +\Delta t \cdot R\left[t^n, U\left(t^n\right)\right]$ & $U = U + U\%t(t=t) * Dt$\\
    \hline
    TVD/SSP Runge-Kutta & $\begin{matrix}
  K_s = R\left(t^n + \gamma_s \Delta t, U^n+\Delta t\sum\limits_{l=1}^{s-1}{\alpha_{s,l} K_l} \right) \\
  U^{n+1} = U^{n} + \Delta t \cdot\sum\limits_{s=1}^{N_{s}}{\beta_s K_s}
\end{matrix}$ & $\begin{matrix}
  stage(s) = stage(s) + stage(ss) * (Dt * self\%alph(s, ss)) \\
  U = U +  stage(s) * (Dt * self\%beta(s))
\end{matrix}$\\
    \hline
    low storage Runge-Kutta & $\begin{matrix}
  K_2 = A_s K_2 + \Delta t \cdot R\left(t^n + C_s \Delta t, K_1\right) \\
  K_1 = K_1 + B_s K_2
\end{matrix}$ & $\begin{matrix}
  stage(2) = stage(2) * self\%A(s) + \\ + stage(1)\%t(t=t + self\%C(s) * Dt) * Dt \\
  stage(1) = stage(1) + stage(2) * self\%B(s)
\end{matrix}$\\
    \hline
  explicit Adams-Bashforth & $\begin{matrix}U\left(t^{n+N_s}\right) = U\left(t^{n+N_s-1}\right) + \\ +\Delta t \sum_{s=1}^{n+N_s}{ b_s \cdot R\left[t^{n+s-1}, U\left(t^{n+s-1}\right)\right]}\end{matrix}$ & $U = U + U\%t(n=s, t=t(s)) * (Dt * self\%b(s))$\\
    \hline
    explicit Leapfrog & $ U\left(t^{n+2}\right) = U\left(t^{n}\right) + 2\Delta t \cdot R\left[t^{n+1}, U\left(t^{n+1}\right)\right]$ & $U = U\%previous\_step(n=1) + U\%t(n=2, t=t) * (Dt * 2.)$\\

  \end{tabular}}
\end{table}

@@

Let us consider the \emph{oscillation} problem, it being a simple, yet interesting IVP. Briefly, the oscillation problem is a prototype problem of non dissipative, oscillatory phenomena. For example, let us consider a pendulum subjected to the Coriolis accelerations without dissipation, the motion equations of which can be described by the ODE system \ref{eq:oscillation}.

\begin{equation}
\begin{matrix}
U_t = R(U)  \\
U = \begin{bmatrix}
x \\
y
\end{bmatrix}\;\;\;
R(U) = \begin{bmatrix}
-f y \\
f x
\end{bmatrix}
\end{matrix}
\label{eq:oscillation}
\end{equation}
where the frequency is chosen as $f=10^4$. The ODE system \ref{eq:oscillation} is completed by the following initial conditions:

\begin{equation}
\begin{matrix}
  x(t_0) = 0 \\
  y(t_0) = 1
\end{matrix}
\label{eq:oscillation-ic}
\end{equation}
where $t_0=0$ is the initial time considered.

The IVP constituted by equations \ref{eq:oscillation} and \ref{eq:oscillation-ic} is (apparently) simple and its exact solution is known:
